---
output: html_document
params:
  version: 1.1
---

# Functions, Best-coding Practices, and Debugging Examples {-}

## Functions

### Example 1

Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need?

```{r}
mean(is.na(x))
mean(is.na(y))
mean(is.na(z))

x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)

round(x / sum(x, na.rm = TRUE) * 100, 1)
round(y / sum(y, na.rm = TRUE) * 100, 1)
round(z / sum(z, na.rm = TRUE) * 100, 1)
```

```{r, error=TRUE}
percent_total <- function(xvec) {
  mean(is.na(x))
  x / sum(x, na.rm = TRUE)
  round(x / sum(x, na.rm = TRUE) * 100, 1)
}

percent_total(xvec = 1)
```

Note that clicking show traceback will allow you to see where the error occured.

```{r, error=TRUE}
percent_total <- function(xvec) {
  mean(is.na(xvec)) # % of values that are NA
  xvec / sum(xvec, na.rm = TRUE)
  return(round(xvec / sum(xvec, na.rm = TRUE) * 100, 1))
}

percent_total(xvec = c(1, NA, 1))

mean(is.na(c(1, NA, 1)))
```

```{r}
library(tidyverse)
percent_total <- function(xvec) {
  tibble(# % of values that are NA
    mean(is.na(xvec)),
    # fraction of total for each value
    xvec / sum(xvec, na.rm = TRUE),
    # percent of total to 1 decimal
    round(xvec / sum(xvec, na.rm = TRUE) * 100, 1))
}

percent_total(xvec = c(1, NA, 1))
```


### Example 2

-   Create a function `LW()` with three arguments: a vector lengths, and values a and b

-   It returns the weights of fishes using the length-weight equation: $W = aL^b$

-   Use the function to calculate the weight (in g) of fish of length 100, 200, 300 cm for:

    -   Mola mola, a = 0.0454, b = 3.05

    -   Regalecus glesne, a = 0.0039, b = 2.90
    
```{r}
#' Title
#'
#' @param lengths in centimeters
#' @param a constant
#' @param b exponent
#'
#' @return weight in grams
#' @export
#'
#' @examples
#' 
LW <- function(lengths, a, b){
  return(a * lengths ^ b)
}

LW(lengths = c(100, 200, 300), a = 0.0454, b = 3.05)

```

    
## Best coding practices

### Code from inside out, running the smallest bits of code possible

### Use RStudio to your advantage

-   Tab-complete
-   Reformat code
-   Reindent lines
-   Rainbow parentheses


## Debugging

-   Missing parentheses/brackets

```{r}
mean(x
```

-   Missing quotes

```{r}
print("Hello)
```

-   Misplaced, missing commas

```{r}
x <- c("1" "3", "7", "Emma")

is.numeric(x)
```

-   Misspelled object names

```{r}

```


### Also use RStudio

-   Debug menu is helpful for writing more advanced functions




