---
output: html_document
params:
  version: 1.0
---

# Interpolation and Extrapolation Examples {.unnumbered}

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(PolynomF)
library(pracma)
```

## Example - CO~2~ at Mana Loa Observatory

The Mana Loa observatory contains the longest *in situ* record of CO~2~ in the atmosphere. We've downloaded this data from  ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_annmean_mlo.txt. 

Let's read in this data and do some extrapolation. Instead of using the `lin_interp` function we wrote, we can just make linear regression model of the data. 

```{r}
coln <- c("year","co2","y")
dfc <- read_delim("data/co2_annmean_mlo.txt",delim = "   ", comment = "#",col_names = coln)

dfc$year <- as.numeric(dfc$year)
dfc$co2 <- as.numeric((dfc$co2))
plot(dfc$year,dfc$co2)


```

Using this data let's extrapolate the CO~2~ level at 2100. We'll start with linear interpolation and in the assignment you will try polynomial and splines.

```{r}
m <- lm(co2 ~ year, data = dfc)

xx <- seq(1970, 2100, 1)
xxd <- data.frame(xx)
colnames(xxd) <- "year"
yy <- predict(m, xxd)

lm_co2 <- data.frame(cbind(xx, yy))

pp <- ggplot() +
  geom_point(
    data = dfc,
    aes(x = year, y = co2),
    size = 1,
    col = 'blue'
  )
pp  +
  geom_line(data = lm_co2, aes(x = xx, y = yy))
```

Well over 550 ppm CO~2~ by 2100. But this doesn't really fit the data very well. Perhaps we could improve a bit by using our `lin_interp` function on the last 2 datapoints. 

```{r}
lin_extrap_data <- tail(dfc, n = 2)
lin_interp(fx1 = lin_extrap_data[[1,"co2"]],
           fx2 = lin_extrap_data[[2,"co2"]], 
           x1 = lin_extrap_data[[1,"year"]],
           x2 = lin_extrap_data[[2,"year"]], 
           x3 = 2100)
m2 <- lm(co2 ~ year, data = lin_extrap_data)

```

620 ppm! That's more than 10% higher, but does it visually fit better? We can use the `geom_abline` function from `ggplot2` to plot a line with the right slope and intercept, although it would probably be easier to go back and make a linear model with the `lin_extrap_data` and use `predict` to make a line as we did above. 

```{r}
pp +
  geom_abline(slope = (lin_extrap_data[[2,"co2"]] - 
                         lin_extrap_data[[1,"co2"]]) / 
                      (lin_extrap_data[[2,"year"]] - 
                        lin_extrap_data[[1,"year"]]), 
              intercept = lin_extrap_data[[1,"co2"]] -
                          (lin_extrap_data[[2,"co2"]] - 
                         lin_extrap_data[[1,"co2"]]) / 
                      (lin_extrap_data[[2,"year"]] - 
                        lin_extrap_data[[1,"year"]])*
  lin_extrap_data[[1,"year"]], 
  color = "grey") + 
  ylim(300, 800) + 
  xlim(1970, 2100)
```

What do you think the predictions from polynomial and splines will be? 

```{r}
co2_poly <- poly_calc(dfc$year, dfc$co2)
dfc_pred <- data.frame(year = xx, co2_poly = co2_poly(xx))
pp + geom_line(data = dfc_pred, mapping = aes(x = year, y = co2_poly))


```

```{r}
m3 <- lm(co2 ~ poly(x = year, degree = 6, raw = TRUE), data = dfc)
summary(m3)
dfc_pred$co2_m3 <- 
  m3 %>% predict(dfc_pred) 
pp + geom_line(data = dfc_pred, mapping = aes(x = year, y = co2_m3))

```
To see the predicted CO~2~ concentration in the year 2100 we can use `tail` to look at the last line of `dfc_pred`
```{r}
dfc_pred %>% tail(n = 1)
```

We could also use `stat_smooth` to plot the model object and include confidence intervals around our prediction. Within `stat_smooth` our formula needs to be in terms of x and y, which correspond to the variables mapped back in the `ggplot` call when we defined `pp`; so looking back `aes(x = year, y = co2)`. To expand the range of the model predictions in `stat_smooth` we need to add `fullrange = TRUE` and then use `xlim` to expand the x-axis. 

```{r}
## Note that previously (and still above for pp) the data and mapping arguments
## were in the `geom_point`, 
## which was why stat_smooth was not working in class.  
# TODO fix pp above, and also rename...
ggplot(data = dfc,
    aes(x = year, y = co2)) +
  geom_point(size = 1) + 
  stat_smooth(method = lm, formula = y ~ poly(x = x, degree = 3, raw = TRUE), 
              fullrange = TRUE) + 
  xlim(1959, 2100)
```

Finally, we can use the method of splines by similarly defining a splines model using the `bs` (stands for [b-splines, which are a type of spline](https://en.wikipedia.org/wiki/B-spline)) from the `splines` package.  

```{r}
library(splines)
m_splines <- lm(co2 ~ splines::bs(year, knots = dfc$co2), data = dfc)
dfc_pred$co2_m_splines <- 
  m_splines %>% predict(dfc_pred) 
pp + geom_line(data = dfc_pred, mapping = aes(x = year, y = co2_m_splines))
```

```{r}
dfc_pred %>% tail(n=1)
```

```{r}
ggplot(data = dfc,
    aes(x = year, y = co2)) +
  geom_point(size = 1) + 
  stat_smooth(method = lm, formula = y ~ bs(x = x, knots = dfc$co2), 
              fullrange = TRUE) + 
  xlim(1959, 2100)
```


